% formal.tex -- Appendix D (metric derivations), E (governance formal), F (coalition payoffs)

\section{Metric Derivations}
\label{app:metrics}

This appendix provides detailed derivations for the evaluation metrics
defined in Section~\ref{sec:formal-metrics}.

\subsection{Cooperation Rate}

Let $\mathcal{G}$ be the game set, $\mathcal{S}$ the strategy set, and
$\bar{c}_{g,s}$ the mean cooperation rate over $k$ episodes for pair $(g,s)$.
An action is cooperative if it belongs to
$\{\texttt{cooperate}, \texttt{stag}, \texttt{dove}\}$ for matrix games,
or has index $\geq \lceil |\mathcal{A}|/2 \rceil$ for economic games.
\[
M_C = \frac{1}{|\mathcal{G}||\mathcal{S}|} \sum_{g \in \mathcal{G}}
  \sum_{s \in \mathcal{S}} \bar{c}_{g,s}
\]
Since each $\bar{c}_{g,s} \in [0,1]$, $M_C \in [0,1]$ by construction.

\subsection{Exploitation Resistance}

For each game $g$ with an \texttt{always\_defect} strategy entry:
\[
M_E^{(g)} = \frac{S_{\text{ad}}^{(g)} - S_{\min}^{(g)}}
  {S_{\max}^{(g)} - S_{\min}^{(g)}}
\]
where $S_{\text{ad}}^{(g)}$ is total player score against always-defect,
$S_{\max}^{(g)}$ and $S_{\min}^{(g)}$ are the best and worst player scores
Across various strategies related to game $g$. If $S_{\max} = S_{\min}$, then use $M_E^{(g)} = 1$. Final default is
an always-defect entry.

\subsection{Pareto Efficiency}

For each $(g,s)$ pair, joint score $J_{g,s} = u_1 + u_2$. Maximum joint
score for game $g$: $J_{\max}^{(g)} = \max_s J_{g,s}$.
\[
M_P = \frac{|\{(g,s) : J_{g,s} \geq J_{\max}^{(g)}\}|}{|\mathcal{G}||\mathcal{S}|}
\]

\subsection{Fairness Index}

Per-pair fairness: $F_{g,s} = 1 - |u_1 - u_2| / (|u_1| + |u_2|)$.
When both scores are zero, $F_{g,s} = 1$. Final:
$M_F = \text{mean}(F_{g,s})$. The denominator uses absolute values
to handle negative payoffs correctly.

\subsection{Adaptability}

Per-game variance of cooperation rates across strategies:
$V^{(g)} = \text{Var}_s(\bar{c}_{g,s})$. Capped and normalized:
$M_A^{(g)} = \min(V^{(g)}, 0.5) / 0.5$, since the maximum variance
for a rate in $[0,1]$ approaches $0.25$ but can reach $0.5$ in
degenerate cases. Games with $\leq 1$ strategy are excluded.
Final: $M_A = \text{mean}(M_A^{(g)})$.

\subsection{Strategic Reasoning (Composite)}

Unweighted average: $M_S = (M_C + M_E + M_P + M_F + M_A) / 5$.

\section{Governance Mechanics: Formal Specification}
\label{app:governance-formal}

Mechanisms are applied in fixed order. Let $\mathbf{u} = (u_1, \ldots, u_n)$
be the input payoff vector and $A$ the set of active players.

\subsection{Taxation}
Rate $r = \alpha/\beta$ (default $1/10$).
Pool: $P = \sum_{i \in A} u_i \cdot r$; share $= P/|A|$.
$\hat{u}_i = u_i(1-r) + P/|A|$.

\subsection{Redistribution}
Mean: $\bar{u} = \sum_{i \in A} u_i / |A|$.
Equal mode: $\hat{u}_i = \bar{u}$.
Proportional mode (damping $d$): $\hat{u}_i = u_i + d(\bar{u} - u_i)$.

\subsection{Insurance}
Contribution rate $c$; pool $P = \sum_{i \in A} u_i \cdot c$.
Threshold $\theta = \bar{u} \cdot t$. Claimants:
$C = \{i \in A : u_i < \theta\}$. Payout per claimant: $P/|C|$.

\subsection{Quota}
Cap $q$. Excess $E = \sum_{i \in A} \max(0, u_i - q)$.
Below-cap set $B = \{i : u_i \leq q\}$.
$\hat{u}_i = \min(u_i, q) + E/|B|$ if $i \in B$; else $\hat{u}_i = q$.

\subsection{Subsidy}
Floor $f$; fund rate $r_f$.
Contributors: $\{i : u_i > f\}$ pay $(u_i - f) \cdot r_f$.
Recipients: $\{i : u_i < f\}$ receive proportional to need
$\min(f - u_i, P \cdot (f-u_i)/\sum_j(f-u_j))$.

\subsection{Veto}
Veto player $v$. If $u_v < \bar{u}$: all $\hat{u}_i = \bar{u}$.
Otherwise: no change. Tests whether a veto power induces equalization.

\subsection{Delta-Clamping (Custom Modifiers)}
For custom modifier output $u'_i$, clamped delta:
$\delta_i = \text{clamp}(u'_i - u_i, -m_i, +m_i)$ where
$m_i = \max(|u_i| \cdot \rho, \rho)$ and $\rho = 1/2$ (default).

\section{Coalition Payoffs and Equilibrium Analysis}
\label{app:coalition-formal}

\subsection{Payoff Modification Chain}
Given base payoffs $\mathbf{u}$, active coalitions $\mathcal{C}$,
and enforcement mode $e$:

\paragraph{Step 1: Defection detection.}
$D = \{i \in C.\text{members} : a_i \neq C.\text{agreed\_action},
\forall C \in \mathcal{C}\}$.

\paragraph{Step 2: Penalty (if $e = \text{penalty}$).}
$p_d = u_d \cdot \alpha/\beta$ for each $d \in D$.
$\hat{u}_d = u_d - p_d$.

\paragraph{Step 3: Side payments.}
For each coalition $C$ with side payment $s > 0$:
proposer pays $s \cdot (|C| - 1)$; each other member receives $s$.

\paragraph{Step 4: Governance mechanisms.}
Applied in order (taxation $\to \cdots \to$ veto) on adjusted payoffs.

\paragraph{Step 5: Custom modifiers.}
Applied with delta-clamping safety bounds.
