% implementation.tex -- Appendix G (dynamic API), H (hyperparams), I (prompts)

\section{Dynamic Game API Reference}
\label{app:api}

The dynamic game creation system (Section~\ref{sec:game-api}) exposes
three construction modes via the \texttt{GameConfig} dataclass:

\begin{description}[nosep]
\item[Matrix.] Specify payoff matrix as nested dict
  $\{a_1 \to \{a_2 \to (u_1, u_2)\}\}$. Validated for action count
  consistency and numeric payoffs.
\item[Symmetric.] Provide reduced specification (upper triangle only);
  expanded to full symmetric matrix automatically.
\item[Custom.] Register a \texttt{Callable[[str, str], tuple[float, float]]}
  that computes payoffs from action profiles. Supports arbitrary game logic.
\end{description}

Games are registered in the global \texttt{GAMES} dictionary with a unique
string key. Extension modules (across \texttt{games\_ext/},
\texttt{games\_info/}, \texttt{games\_market/}, \texttt{games\_coop/},
\texttt{games\_meta/}, \texttt{games\_adaptive/}) are loaded dynamically
via \texttt{\_load\_extensions()}. Adaptive games are registered as
factories in \texttt{GAME\_FACTORIES}, producing fresh \texttt{GameConfig}
instances with mutable state on each instantiation.

\section{Hyperparameters}
\label{app:hyperparams}

\subsection{Training Hyperparameters}

\begin{table}[h]
\centering\small
\caption{GRPO training configuration.}
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Base models & 8 open-weight (1B--27B) + 4 API baselines \\
Group size ($G$) & 8 \\
Max completion length & 16 tokens \\
Per-device batch size & 4 \\
Gradient accumulation & 4 \\
Effective batch size & $4 \times 4 = 16$ \\
Learning rate & $5 \times 10^{-6}$ \\
Max training steps & 500 \\
Dataset size & 1000 prompts \\
Precision & bf16 (H100) \\
Invalid move penalty & $-2.0$ \\
Environment error default & $0.0$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Evaluation Hyperparameters}

\begin{table}[h]
\centering\small
\caption{Tournament and evaluation configuration.}
\begin{tabular}{@{}lr@{}}
\toprule
\textbf{Parameter} & \textbf{Value} \\
\midrule
Episodes per (game, strategy) & 3 \\
Train/eval split ratio & 78\% / 22\% \\
Split seed & 42 \\
Min eval domain fraction & 20\% \\
Cooperation rate cap (adaptability) & 0.5 \\
Composite metric weight (each) & 1/5 \\
\bottomrule
\end{tabular}
\end{table}

\section{Prompt Templates}
\label{app:prompts}

The system prompt for game-theoretic evaluation:

\begin{quote}\small\ttfamily
You are an expert game theory player. You will be given the current state
of a 2-player strategic game and must choose your move to maximize your
long-term cumulative payoff.

Rules:\\
- Read the game description carefully\\
- Consider your opponent's strategy and history\\
- Respond with ONLY the move name, nothing else\\
- Your response must be exactly one of the available moves listed
\end{quote}

Each user prompt includes: game name, description, available moves,
opponent strategy, round number (e.g., ``Round 3/10''), cumulative score,
and full action history formatted as
``Round $k$: you=X, opponent=Y, your payoff=$\pm z$''.
