% eval_protocol.tex -- Section 10: Evaluation Protocol

\section{Evaluation Protocol}
\label{sec:eval-protocol}

This section describes the comprehensive evaluation protocol used to assess
agent behavior across the full Kant game library, including strategies,
tournament structure, formal metrics, and stratified game splitting.

Evaluation measures whether game-theoretic training produces agents that are
not only strategically competent but also cooperative, fair, and robust to
exploitation---properties that transfer to broader alignment.

\subsection{Strategy Library}
\label{sec:strategy-library}

The evaluation protocol employs 24 opponent strategies spanning four categories.

\begin{table}[h]
\centering\small
\caption{Opponent strategy library. Base strategies (11) apply to all matrix games;
economic strategies (6) are game-specific; N-player (5), coalition (4), and governance
(4) strategies extend to multi-agent settings.}
\label{tab:strategies}
\begin{tabular}{@{}llp{5.8cm}@{}}
\toprule
\textbf{Category} & \textbf{Strategy} & \textbf{Logic} \\
\midrule
\multirow{5}{*}{Unconditional}
& Always Cooperate & Always plays action$[0]$ \\
& Always Defect & Always plays action$[1]$ \\
& Random & Uniform random over actions \\
& Mixed & Cooperate w.p.\ $0.7$, defect w.p.\ $0.3$ \\
\midrule
\multirow{5}{*}{Reciprocal}
& Tit-for-Tat & Cooperate first; then mirror opponent \\
& Tit-for-Two-Tats & Defect only after 2 consecutive opponent defections \\
& Grudger & Cooperate until first defection, then always defect \\
& Pavlov & Repeat if same action chosen; else switch \\
& Suspicious TFT & Defect first; then mirror opponent \\
& Generous TFT & Like TFT but cooperate 90\% after defection \\
& Adaptive & Cooperate if opponent cooperated $>50\%$ historically \\
\midrule
Economic & Ultimatum Fair/Low & Offer $5$ or $3$ of pot $10$ \\
& Trust Fair/Generous & Invest max; return $\tfrac{1}{3}$ or $\tfrac{1}{2}$ \\
& PG Fair/Free-Rider & Contribute $50\%$ or $10\%$ of endowment \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Tournament Structure}
\label{sec:tournament-structure}

Evaluation proceeds via round-robin tournament. For each
$(\text{game}, \text{strategy})$ pair, the \texttt{TournamentRunner}
executes $k$ episodes (default $k=3$) and aggregates results:

\begin{enumerate}[nosep]
    \item For each game $g \in \mathcal{G}$ and strategy $s \in \mathcal{S}$:
    \item \quad For each episode $e \in \{1, \ldots, k\}$:
    \item \quad\quad Reset environment with $(g, s)$; play until terminal state
    \item \quad\quad Record per-round actions, payoffs, cooperation flags
    \item \quad Aggregate: total scores, mean cooperation rate over $k$ episodes
    \item Compute six metrics over the full results matrix (Section~\ref{sec:formal-metrics})
\end{enumerate}

\noindent With $88$ games and $11$ base strategies, a full tournament
executes $88 \times 11 \times 3 = 2{,}904$ episodes.

\subsection{Formal Metrics}
\label{sec:formal-metrics}

We define six metrics, each normalized to $[0, 1]$:

\paragraph{Cooperation Rate ($M_C$).}
Mean fraction of cooperative actions across all $(g, s)$ pairs:
$M_C = \frac{1}{|\mathcal{G}||\mathcal{S}|} \sum_{g,s} \bar{c}_{g,s}$,
where $\bar{c}_{g,s}$ is the cooperation rate averaged over $k$ episodes.
Cooperative actions are $\{\texttt{cooperate}, \texttt{stag}, \texttt{dove}\}$
for matrix games; actions at or above the median index for economic games.

\paragraph{Exploitation Resistance ($M_E$).}
Per-game performance against \texttt{always\_defect} relative to
best/worst:
$M_E^{(g)} = \frac{S_{\text{ad}}^{(g)} - S_{\min}^{(g)}}{S_{\max}^{(g)} - S_{\min}^{(g)}}$,
averaged over games. Score of $1.0$ means no payoff loss to exploitation.

\paragraph{Pareto Efficiency ($M_P$).}
Fraction of $(g,s)$ pairs where joint payoff $u_1 + u_2$ equals the
maximum joint payoff observed for that game:
$M_P = \frac{|\{(g,s) : J_{g,s} \geq J_{\max}^{(g)}\}|}{|\mathcal{G}||\mathcal{S}|}$.

\paragraph{Fairness Index ($M_F$).}
Inverted normalized payoff difference:
$M_F^{(g,s)} = 1 - \frac{|u_1 - u_2|}{|u_1| + |u_2|}$, averaged over
all pairs. Score of $1.0$ indicates perfectly equal payoffs.

\paragraph{Adaptability ($M_A$).}
Normalized variance of cooperation rate across strategies within each game:
$M_A^{(g)} = \frac{\min(\text{Var}_s(\bar{c}_{g,s}),\; 0.5)}{0.5}$,
averaged over games. High variance indicates context-dependent behavior.

\paragraph{Strategic Reasoning ($M_S$).}
Unweighted composite: $M_S = \frac{1}{5}(M_C + M_E + M_P + M_F + M_A)$.

\subsection{Stratified Game Splitting}
\label{sec:splitting}

To ensure evaluation generalization, we introduce domain-aware stratified
splitting with seed $42$ for reproducibility:

\begin{itemize}[nosep]
    \item \textbf{Train set}: $78\%$ of games ($\approx 69$ games)
    \item \textbf{Eval set}: $22\%$ of games ($\approx 19$ games)
    \item \textbf{Domain constraint}: each domain tag has $\geq 20\%$
          representation in the eval set
\end{itemize}

\noindent The domain constraint ensures no single game category (e.g.,
classical dilemmas, market competition) is entirely absent from evaluation.
Combined with procedurally generated games (Section~\ref{sec:procedural}),
this tests whether strategic reasoning transfers to unseen game structures.
