% safety.tex -- Section 12: Safety Transfer Evaluation

\section{Safety Transfer Evaluation}
\label{sec:safety}

A central hypothesis of the Kant project is that game-theoretic training
transfers to improved performance on established AI safety benchmarks. This
section describes the external benchmark suite, experimental setup, and
preliminary results.

% TODO: Expand hypothesis motivation

\subsection{Hypothesis: Game Theory to Alignment}
\label{sec:hypothesis}

We hypothesize that training language models to reason about cooperation,
fairness, trust, and strategic interaction in game-theoretic settings
develops internal representations that transfer to broader alignment tasks.
Specifically, an agent trained on Kant's game library should exhibit:

\begin{itemize}[nosep]
    \item Improved refusal of harmful requests (HarmBench)
    \item Better ethical reasoning (ETHICS)
    \item More truthful responses (TruthfulQA)
    \item Appropriate compliance with safe prompts (XSTest)
    \item Maintained general capability (MT-Bench)
\end{itemize}

% TODO: Formalize transfer hypothesis

\subsection{External Benchmark Suite}
\label{sec:external-benchmarks}

We evaluate on five established benchmarks:

\begin{description}[nosep]
    \item[HarmBench~\citep{mazeika2024harmbench}.] Tests refusal of harmful
        requests across multiple attack categories.
    \item[ETHICS~\citep{hendrycks2021ethics}.] Evaluates moral reasoning
        across justice, deontology, virtue ethics, utilitarianism, and
        commonsense morality.
    \item[TruthfulQA~\citep{lin2022truthfulqa}.] Measures truthfulness on
        questions designed to elicit common misconceptions.
    \item[XSTest~\citep{rottger2024xstest}.] Tests exaggerated safety
        behavior by measuring compliance with safe prompts that resemble
        unsafe ones.
    \item[MT-Bench~\citep{zheng2023judging}.] Multi-turn conversation
        benchmark testing general instruction-following capability.
\end{description}

% TODO: Benchmark configuration details

\subsection{Experimental Setup}
\label{sec:safety-setup}

We compare three model configurations:

\begin{enumerate}[nosep]
    \item \textbf{Base model}: unmodified pretrained language model
    \item \textbf{Kant-GRPO}: base model fine-tuned with GRPO on game environments
    \item \textbf{Kant-DPO}: base model fine-tuned with DPO on game trajectory pairs
\end{enumerate}

% TODO: Model specifications, training details, evaluation protocol

\subsection{Results}
\label{sec:safety-results}

% TODO: Results table comparing base, Kant-GRPO, and Kant-DPO across all benchmarks
% TODO: Per-category breakdowns
% TODO: Statistical significance tests

\textit{Results are pending completion of training runs. See
Appendix~\ref{app:results-ext} for extended result tables.}
