% training.tex -- Section 11: Training Pipeline

\section{Training Pipeline}
\label{sec:training}

Kant includes a training pipeline that uses game-theoretic environments to
improve language model alignment through reinforcement learning. This section
describes the GRPO and DPO training methods, curriculum learning strategy,
and reward shaping approach.

% TODO: Introductory paragraph connecting training to alignment hypothesis

\subsection{GRPO on Game Environments}
\label{sec:grpo}

Group Relative Policy Optimization (GRPO)~\citep{shao2024deepseekmath}
generates multiple trajectory rollouts per game episode and optimizes the
policy by comparing relative rewards within each group. Applied to Kant's
game environments, GRPO trains agents to discover cooperative equilibria
through direct environment interaction.

% TODO: GRPO algorithm adapted for game environments
% TODO: Reward function specification

\subsection{DPO via Trajectory Preference Pairs}
\label{sec:dpo}

Direct Preference Optimization (DPO)~\citep{rafailov2023direct} trains on
preference pairs constructed from game trajectories. For each game episode,
we generate a ``preferred'' trajectory (cooperative, fair, strategically
sound) and a ``rejected'' trajectory (exploitative, unfair, or naive). The
DPO loss directly optimizes the policy to prefer aligned behavior without
requiring an explicit reward model.

% TODO: Trajectory pair construction algorithm
% TODO: DPO loss formulation for game trajectories

\subsection{Curriculum Learning}
\label{sec:curriculum}

Training proceeds through a curriculum that progressively expands the game
domain:

\begin{enumerate}[nosep]
    \item \textbf{Phase~I}: Classical dilemmas (PD, Stag Hunt, Hawk-Dove)
    \item \textbf{Phase~II}: Extended matrix and sequential games
    \item \textbf{Phase~III}: N-player and coalition games
    \item \textbf{Phase~IV}: Governance-enabled games and dynamic creation
\end{enumerate}

% TODO: Phase transition criteria, game sampling strategy

\subsection{Reward Shaping}
\label{sec:reward-shaping}

Raw game payoffs are transformed via a reward shaping function that balances
individual performance with cooperative and fairness objectives. The shaped
reward incorporates:

\begin{itemize}[nosep]
    \item Individual payoff (self-interest signal)
    \item Joint payoff (cooperation signal)
    \item Payoff difference penalty (fairness signal)
    \item Exploitation penalty (safety signal)
\end{itemize}

% TODO: Reward shaping formula, weight hyperparameters
