% governance.tex -- Section 8: Meta-Governance Engine

\section{Meta-Governance Engine}
\label{sec:governance}

The meta-governance engine is Kant's most distinctive architectural feature:
it allows players to \emph{change the rules of the game during play}. By
proposing, voting on, and applying rule modifications, agents participate in
a form of constitutional self-governance that directly parallels Constitutional
AI~\citep{bai2022constitutional} but operates through democratic mechanism
design rather than static principles.

% TODO: Expand connection to Constitutional AI

\subsection{Constitutional Mechanics}
\label{sec:constitutional}

The governance system operates as a layer atop any N-player game. Between
action rounds, a governance phase allows players to propose modifications to
the game's rules, vote on proposals, and apply approved changes. This creates
a two-level game: the \emph{object game} (the underlying strategic interaction)
and the \emph{meta-game} (the governance process itself).

% TODO: Formal two-level game model

\subsection{Proposal System}
\label{sec:proposals}

Three types of proposals are supported:

\begin{description}[nosep]
    \item[Parameter Proposals.] Modify numeric game parameters (e.g., payoff
        values, contribution multipliers, round counts) within pre-defined
        bounds.
    \item[Mechanic Proposals.] Activate or deactivate one of six composable
        mechanisms (Section~\ref{sec:mechanisms}).
    \item[Custom Proposals.] Free-form modifications specified as payoff
        delta functions, subject to delta-clamping safety constraints
        (Section~\ref{sec:delta-clamping}).
\end{description}

% TODO: Proposal format specification, example proposals

\subsection{Democratic Voting}
\label{sec:voting}

Proposals are decided by majority vote with a configurable threshold (default:
simple majority). Each player casts one vote per proposal. Voting is
simultaneous and secret by default, though open voting modes are supported.
Approved proposals take effect in the next round.

% TODO: Voting protocol formal specification

\subsection{Composable Mechanisms}
\label{sec:mechanisms}

Six governance mechanisms can be proposed, voted on, and composed:

\begin{enumerate}[nosep]
    \item \textbf{Taxation} --- a fraction of each player's payoff is collected
          into a governance pool.
    \item \textbf{Redistribution} --- the governance pool is distributed
          equally among all players (or according to a voted-on formula).
    \item \textbf{Insurance} --- players pay premiums to a pool that
          compensates low-scoring players, reducing variance.
    \item \textbf{Quota} --- minimum or maximum action constraints (e.g.,
          minimum contribution in Public Goods).
    \item \textbf{Subsidy} --- bonus payments for specific actions (e.g.,
          cooperation bonuses).
    \item \textbf{Veto} --- designated players or coalitions can block
          specific actions or proposals.
\end{enumerate}

% TODO: Formal mechanism specifications with payoff transformations

\subsection{Custom Modifiers with Delta-Clamping Safety}
\label{sec:delta-clamping}

Custom proposals specify arbitrary payoff modifications via delta functions.
To prevent destabilizing changes, a delta-clamping mechanism bounds the
magnitude of any single modification.

% TODO: Delta-clamping formula, safety bounds, worked example

\subsection{Governance-Coalition Interaction Loop}
\label{sec:gov-coalition}

When coalition formation and governance are active simultaneously, a rich
interaction loop emerges: coalitions may coordinate voting blocs, propose
self-serving rules, or use governance to enforce coalition agreements. This
creates emergent institutional dynamics analogous to real-world political
systems.

% TODO: Interaction diagram, emergent behavior examples

\subsection{Connection to Constitutional AI}
\label{sec:constitutional-ai}

The governance engine operationalizes key ideas from Constitutional
AI~\citep{bai2022constitutional} in a multi-agent game-theoretic setting.
Rather than encoding principles as static rules, Kant allows principles to
emerge through democratic deliberation. This tests whether agents can
\emph{construct} fair governance systems, not merely follow pre-specified ones.

% TODO: Detailed comparison table: Constitutional AI vs. Kant governance
