% library.tex -- Section 5: Game Library (114 games across 10 domains)

\section{Game Library}
\label{sec:game-library}

The Kant framework provides a library of 114 games spanning 10 strategic domains.
Each game is implemented as a \texttt{GameConfig} with formally specified payoff
structures, action spaces, and metadata tags. This section catalogues every game
by domain; Appendix~\ref{app:catalog} provides the full specification table.

% Add introductory paragraph connecting game diversity to alignment coverage

\subsection{Classical Dilemmas}
\label{sec:classical}

The foundational two-player simultaneous-move games that define the core tensions
in strategic interaction: cooperation vs.\ self-interest, coordination under risk,
conflict escalation, and zero-sum competition.

\begin{table}[h]
\centering\small
\caption{Classical dilemma payoff matrices $(u_1, u_2)$.}
\label{tab:classical}
\begin{tabular}{@{}ll cc@{}}
\toprule
\textbf{Game} & & \multicolumn{2}{c}{\textbf{Player 2}} \\
\midrule
\textbf{Prisoner's Dilemma} & & Cooperate & Defect \\
& Cooperate & $(3,3)$ & $(0,5)$ \\
& Defect & $(5,0)$ & $(1,1)$ \\
\midrule
\textbf{Stag Hunt} & & Stag & Hare \\
& Stag & $(4,4)$ & $(0,3)$ \\
& Hare & $(3,0)$ & $(2,2)$ \\
\midrule
\textbf{Hawk-Dove} & & Hawk & Dove \\
& Hawk & $(-1,-1)$ & $(3,1)$ \\
& Dove & $(1,3)$ & $(2,2)$ \\
\midrule
\textbf{Matching Pennies} & & Heads & Tails \\
& Heads & $(1,-1)$ & $(-1,1)$ \\
& Tails & $(-1,1)$ & $(1,-1)$ \\
\bottomrule
\end{tabular}
\end{table}

The Prisoner's Dilemma satisfies $T > R > P > S$ ($5 > 3 > 1 > 0$) with unique
Nash equilibrium (Defect, Defect). Stag Hunt has two pure NE---$(S,S)$ is
payoff-dominant while $(H,H)$ is risk-dominant. Hawk-Dove admits a mixed NE at
$p(\text{Hawk}) = \tfrac{v-c}{v} = \tfrac{1}{4}$. Matching Pennies is strictly
competitive with unique mixed NE at $(\tfrac{1}{2}, \tfrac{1}{2})$.

\subsection{Prisoner's Dilemma Variants}
\label{sec:pd-variants}

Extensions to the canonical PD that test nuanced cooperative reasoning under
modified incentive structures.

\begin{itemize}[nosep]
    \item \textbf{Optional PD} --- adds an ``abstain'' action yielding a safe
          intermediate payoff, testing whether agents exploit the exit option.
    \item \textbf{Asymmetric PD} --- unequal payoff rows break symmetry,
          creating differential incentives to cooperate.
    \item \textbf{Donation Game} --- cooperation costs $c$ to self but gives $b > c$
          to the partner, testing pure altruistic transfer.
    \item \textbf{Friend-or-Foe} --- binary ally/enemy classification where
          perceived type determines payoff structure.
    \item \textbf{Peace-War Game} --- geopolitical framing where mutual peace
          yields $(3,3)$, unilateral war yields $(5,0)$, and mutual war $(-1,-1)$.
\end{itemize}

\subsection{Extended Matrix Games}
\label{sec:extended-matrix}

Additional $2 \times 2$ and $3 \times 3$ matrix games covering coordination,
anti-coordination, and cyclic dominance.

\begin{table}[h]
\centering\small
\caption{Extended matrix game payoffs. BoS and Pure Coordination test coordination;
Deadlock and Harmony test dominant-strategy reasoning; RPS tests mixed strategies.}
\label{tab:extended}
\begin{tabular}{@{}lcc|lcc@{}}
\toprule
\textbf{BoS} & Opera & Football & \textbf{Deadlock} & Coop & Defect \\
\midrule
Opera & $(3,2)$ & $(0,0)$ & Coop & $(2,2)$ & $(1,4)$ \\
Football & $(0,0)$ & $(2,3)$ & Defect & $(4,1)$ & $(3,3)$ \\
\midrule
\textbf{Harmony} & Coop & Defect & \textbf{Pure Coord} & Left & Right \\
\midrule
Coop & $(4,4)$ & $(2,3)$ & Left & $(2,2)$ & $(0,0)$ \\
Defect & $(3,2)$ & $(1,1)$ & Right & $(0,0)$ & $(2,2)$ \\
\bottomrule
\end{tabular}
\end{table}

\noindent Rock-Paper-Scissors uses payoffs $\{+1, 0, -1\}$ (win, draw, lose) with
unique mixed NE at uniform $(\tfrac{1}{3}, \tfrac{1}{3}, \tfrac{1}{3})$.
RPSLS extends to five actions with the same cyclic dominance structure.

\subsection{Sequential and Bargaining Games}
\label{sec:sequential-bargaining}

Games with sequential moves or explicit bargaining protocols testing backward
induction, trust, and fairness.

\paragraph{Ultimatum Game.} A proposer splits a pot of $E=10$. The responder
accepts (yielding $(E - \text{offer}, \text{offer})$) or rejects (both get $0$).
The subgame-perfect equilibrium offers $\epsilon$, but fairness norms predict
$\approx 50\%$ splits.

\paragraph{Trust Game.} An investor sends $x \in [0, E]$ ($E=10$); the amount
is tripled ($3x$). The trustee returns $y \in [0, 3x]$. Payoffs:
$(E - x + y,\; 3x - y)$. Backward induction predicts $y=0$, hence $x=0$,
but reciprocity enables mutually beneficial exchange.

\paragraph{Centipede Game.} An alternating take/pass game with initial pot $4$
and growth multiplier $2$. At each of $6$ stages, the active player takes
($75\%$ of pot) or passes (pot doubles). Backward induction predicts immediate
taking; cooperative play grows the pot exponentially.

\paragraph{Stackelberg Competition.} Leader chooses quantity $q_L$, follower
observes and chooses $q_F$. Price $P = 12 - (q_L + q_F)$, marginal cost
$c = 2$. The leader commits to the Stackelberg quantity $q_L^* = 5$,
yielding first-mover advantage.

\noindent Additional games include Dictator ($E=10$, no rejection), Rubinstein
alternating offers, Nash Demand (simultaneous claims), and Double Auction
(bilateral price discovery).

\subsection{Information and Signaling Games}
\label{sec:signaling}

Games testing strategic information transmission and belief updating.
Beer-Quiche tests separating vs.\ pooling equilibria (tough type prefers
beer, payoff $3$ vs.\ $1$). Lemon Market models adverse selection with
quality $q \in \{2, 8\}$ and price $p \in [0,10]$. Cheap Talk tests
informative equilibria under costless communication (aligned payoff $3$,
mismatch $0$). Also includes Spence Signaling and Bayesian Persuasion.

\subsection{Bayesian and Incomplete Information Games}
\label{sec:bayesian}

Games where players have private information about types, states, or values.

\begin{itemize}[nosep]
    \item \textbf{Global Game} --- coordination under noisy signals
    \item \textbf{Jury Voting} --- information aggregation via votes
    \item \textbf{Information Cascade} --- herding and sequential social learning
    \item \textbf{Adverse Selection} --- screening under asymmetric information
\end{itemize}

% Bayesian games test coordination under noisy signals and information aggregation.

\subsection{Communication and Mechanism Design}
\label{sec:mechanism}

Games testing the role of communication channels and mechanism design principles.

\begin{itemize}[nosep]
    \item \textbf{Cheap Talk PD} --- pre-play communication in the Prisoner's Dilemma
    \item \textbf{Binding Commitment} --- enforceable pre-play agreements
    \item \textbf{Correlated Equilibrium} --- coordination via shared signals
    \item \textbf{Focal Point Game} --- coordination via salience (Schelling points)
    \item \textbf{Mediation Game} --- third-party recommendation compliance
\end{itemize}

% Communication protocols and equilibrium refinements

\subsection{Market Competition}
\label{sec:market}

Games modeling competitive market interactions.
Cournot Duopoly uses demand $P = 12 - Q$ with $\text{MC} = 2$;
Bertrand tests price competition (NE at marginal cost).
Traveler's Dilemma (claims $\in [2,100]$, bonus $\pm 2$) exhibits the
paradox of iterated dominance vs.\ cooperative claiming.
Dollar Auction (prize $= 1$, max bid $= 100$) tests escalation commitment.
Minority Game rewards anti-coordination (payoff $5$ for minority, $1$ for
majority). Also includes Hotelling, Entry Deterrence, and Diner's Dilemma.

\subsection{Auctions}
\label{sec:auctions}

Strategic bidding under various mechanisms.
First-Price Sealed-Bid requires bid shading (NE at $b_i = v_i/2$ under
uniform priors). Vickrey (second-price) has truthful bidding as weakly
dominant strategy. All-Pay Auction requires full expenditure regardless of
outcome, testing risk assessment and commitment.

\subsection{Cooperative Game Theory}
\label{sec:cooperative}

Games testing cooperative solution concepts and fair division.
Shapley Value uses $v(\{i\}) = 2$, $v(\{i,j\}) = 8$, $v(N) = 12$; players
claim shares and receive payoff if claims are feasible ($\sum c_i \leq 12$).
Core Allocation divides a pot of $10$ subject to coalition rationality.
Weighted Voting (quota $= 6$, weights $3$ and $4$) tests power index
awareness. Stable Matching uses three-option preferences with payoffs
$\{5, 3, 1\}$ for top/mid/low matches. Also includes Median Voter and
Approval Voting.

\subsection{Contest and Conflict}
\label{sec:contest}

Games modeling resource allocation under competition and conflict.

\begin{itemize}[nosep]
    \item \textbf{Colonel Blotto} --- multi-battlefield resource allocation
    \item \textbf{War of Attrition} --- timing and endurance
    \item \textbf{Tullock Contest} --- probabilistic conflict resolution
    \item \textbf{Inspection Game} --- audit and compliance
    \item \textbf{Security Game} --- attacker-defender resource allocation
    \item \textbf{Network Game} --- strategic interaction on graphs
    \item \textbf{Dueling Game} --- sequential escalation decisions
\end{itemize}

% Tullock contest uses probabilistic success function p_i = x_i / sum(x_j).

\subsection{Principal-Agent and Contracts}
\label{sec:principal-agent}

Games modeling delegation, incentive alignment, and contract design.

\begin{itemize}[nosep]
    \item \textbf{Moral Hazard} --- hidden action after contracting
    \item \textbf{Screening Game} --- menu design for type separation
    \item \textbf{Gift Exchange} --- reciprocity in employment relationships
\end{itemize}

% Tests incentive alignment under delegation and hidden action/information.

\subsection{Behavioral and Dynamic Games}
\label{sec:behavioral}

Games testing bounded rationality, dynamic interaction, and risk assessment.

\begin{itemize}[nosep]
    \item \textbf{Bank Run} --- coordination failure under panic
    \item \textbf{Beauty Contest} --- iterated reasoning about others' beliefs
    \item \textbf{Markov Game} --- state-dependent strategic interaction
    \item \textbf{Stochastic PD} --- probabilistic continuation
    \item \textbf{Risk Dominance Game} --- risk vs.\ payoff dominance selection
    \item \textbf{Threshold Public Good} --- binary provision point
\end{itemize}

% Bank Run and Beauty Contest test bounded rationality; Stochastic PD tests
% discount-rate reasoning; Threshold PG tests binary provision points.

\subsection{Advanced and Evolutionary Games}
\label{sec:advanced}

Specialized games testing evolutionary stability and advanced strategic concepts.

\begin{itemize}[nosep]
    \item \textbf{Preemption Game} --- first-mover advantage timing
    \item \textbf{War of Gifts} --- competitive generosity
    \item \textbf{Penalty Shootout} --- simultaneous mixed strategy under pressure
    \item \textbf{Hawk-Dove-Bourgeois} --- property rights and evolutionary stability
    \item \textbf{Continuous PD} --- cooperation level as continuous choice
    \item \textbf{Discounted PD} --- time-discounted repeated interaction
\end{itemize}

% HD-Bourgeois adds property rights (ESS when ownership respected).
% Continuous PD uses $c \in [0,1]$ cooperation level.

\subsection{Adaptive Games}
\label{sec:adaptive}

Five factory-generated games whose payoff structures evolve based on
interaction history, testing whether agents can reason about non-stationary
incentives.

\begin{itemize}[nosep]
    \item \textbf{Adaptive PD} --- payoff multiplier grows with mutual
          cooperation, shrinks with mutual defection
    \item \textbf{Arms Race} --- Hawk-Dove variant where conflict costs
          escalate on repeated hawk-hawk outcomes
    \item \textbf{Trust Erosion} --- trust multiplier decays with mutual
          defection, recovers with cooperation
    \item \textbf{Market Dynamics} --- Cournot-like duopoly with demand
          curve shifts based on aggregate output history
    \item \textbf{Reputation Payoffs} --- PD with payoff bonus
          proportional to cumulative cooperation rate
\end{itemize}

\noindent Each factory produces fresh \texttt{GameConfig} instances with
mutable state, registered in the \texttt{GAME\_FACTORIES} dictionary.

\subsection{Procedural Generation}
\label{sec:procedural}

Runtime-generated games for open-ended evaluation and generalization testing.
Random Symmetric/Asymmetric/Zero-Sum/Coordination generators produce novel
$2\times 2$ payoff matrices with structural constraints (e.g., $u_1 + u_2 = 0$
for zero-sum, diagonal dominance for coordination). Parameterized Chicken
continuously varies the Hawk-Dove cost/benefit ratio. These generators
enable evaluation on unseen game structures, testing transfer of strategic
reasoning beyond the training distribution (Section~\ref{sec:splitting}).
