% results_ext.tex -- Appendix J (extended results), K (external benchmark details)

\section{Extended Game-Theoretic Results}
\label{app:results-ext}

Extended experimental results with per-game and per-strategy breakdowns.

\textit{Per-game metric breakdowns, per-strategy heatmaps, learning curves,
and statistical significance tests will be populated after full experimental
runs with trained checkpoints. Key analyses will include:}

\begin{itemize}[nosep]
\item Per-game $M_C$, $M_E$, $M_P$, $M_F$, $M_A$ across all 88 games
\item Strategy-game heatmaps showing cooperation rate by opponent type
\item Learning curves across curriculum phases (I--IV)
\item Paired bootstrap confidence intervals for metric comparisons
\item Coalition game cooperation/defection rates by enforcement mode
\item Governance game mechanism adoption frequencies
\end{itemize}

\section{External Benchmark Details}
\label{app:external-details}

Detailed per-category results for each external safety benchmark.

\subsection{HarmBench Results}

Per-attack-category refusal rate comparison: base model vs.\ GRPO-trained
vs.\ DPO-trained. Categories include direct harm, social engineering,
misinformation, and dual-use research.

\textit{Results pending trained checkpoint availability.}

\subsection{ETHICS Results}

Per-category accuracy (justice, deontology, virtue ethics, utilitarianism,
commonsense morality) across model variants.

\textit{Results pending trained checkpoint availability.}

\subsection{TruthfulQA Results}

MC1 (single true answer) and MC2 (multi-true) scores, with per-category
truthfulness rates across 38 categories.

\textit{Results pending trained checkpoint availability.}

\subsection{XSTest Results}

Safe prompt compliance rates and false refusal analysis. XSTest specifically
tests whether safety training causes over-refusal on benign prompts.

\textit{Results pending trained checkpoint availability.}

\subsection{MT-Bench Results}

Per-category instruction-following scores (writing, roleplay, reasoning,
math, coding, extraction, STEM, humanities) and turn-level analysis.

\textit{Results pending trained checkpoint availability.}
