% background.tex -- Related Work and Benchmark Design

\section{Related Work}
\label{sec:related}

\paragraph{MACHIAVELLI.}
Pan et al.~\cite{pan2023machiavelli} introduced a text-based benchmark
measuring trade-offs between reward maximization and ethical behaviour
in choose-your-own-adventure games. Their work highlights the tension
between performance and social norms but relies on rich narrative
environments that conflate language understanding with strategic choice.
\textsc{KantBench} complements this line of work by isolating
strategic reasoning in minimal, formally specified games.

\paragraph{Melting Pot.}
Leibo et al.~\cite{leibo2021meltingpot} proposed Melting Pot for
evaluating multi-agent reinforcement learning via substrate games with
emergent social dynamics. While Melting Pot supports rich multi-agent
scenarios, its substrate complexity can obscure the specific strategic
competencies under test. Our benchmark deliberately restricts the game
space to canonical forms where optimal and equilibrium play are
well-understood.

\paragraph{OpenSpiel.}
Lanctot et al.~\cite{lanctot2019openspiel} provide a broad framework
for game-playing research supporting dozens of game types. OpenSpiel is
designed for algorithm development rather than agent evaluation; it does
not ship pre-defined opponent pools or alignment-oriented metrics.
\textsc{KantBench} occupies a complementary niche: a
purpose-built evaluation harness with standardized opponents and
composite scoring.

\paragraph{Axelrod Library.}
Knight et al.~\cite{knight2016axelrod} offer an open-source framework
for studying the iterated Prisoner's Dilemma, including over 200
strategies and tournament infrastructure. We draw inspiration from this
project but extend the scope to multiple game types and introduce
metrics that go beyond raw tournament scores to capture alignment-relevant
properties such as fairness and exploitation resistance.

\section{Benchmark Design}
\label{sec:design}

\subsection{Game Suite}

\textsc{KantBench} includes six games partitioned into two
families. Table~\ref{tab:payoffs} summarizes the payoff parameters
for each matrix game, and Table~\ref{tab:opponents} lists the
opponent strategies.

\paragraph{Matrix games.}
Prisoner's Dilemma~\cite{rapoport1965prisoner}, Stag
Hunt, and Hawk-Dove are two-player simultaneous-move games played over
multiple rounds (default: ten). Actions and payoffs are defined by a
normal-form matrix; see Figure~\ref{fig:payoff-matrices}.

\paragraph{Sequential games.}
The Ultimatum Game, Trust Game, and Public Goods Game are single-shot
interactions where one or both players choose a continuous allocation
(discretized into integer amounts). These games test fairness norms,
trust calibration, and collective-action incentives, respectively.

\begin{table}[t]
\centering
\caption{Payoff parameters for the three matrix games. Each cell shows
$(u_{\text{row}}, u_{\text{col}})$.}
\label{tab:payoffs}
\small
\begin{tabular}{lcccc}
\toprule
\textbf{Game} & \textbf{(C,C)} & \textbf{(C,D)} & \textbf{(D,C)} & \textbf{(D,D)} \\
\midrule
Prisoner's Dilemma & $(3,3)$ & $(0,5)$ & $(5,0)$ & $(1,1)$ \\
Stag Hunt           & $(4,4)$ & $(0,3)$ & $(3,0)$ & $(2,2)$ \\
Hawk-Dove           & $(-1,-1)$ & $(3,1)$ & $(1,3)$ & $(2,2)$ \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Sequential Game Parameters}

The three sequential games use the following parameters:

\begin{itemize}[nosep]
    \item \textbf{Ultimatum Game}: a fixed pot is split by the proposer;
          the responder accepts or rejects. Rejection yields zero for
          both parties.
    \item \textbf{Trust Game}: an investor sends part of an endowment;
          the amount is tripled and forwarded to a trustee, who decides
          how much to return.
    \item \textbf{Public Goods Game}: each participant contributes from
          a private endowment to a common pool. The pool is multiplied
          by a factor of $\frac{3}{2}$ and split equally among all
          participants.
\end{itemize}

\subsection{Opponent Strategies}

\begin{table}[t]
\centering
\caption{Built-in opponent strategies. Matrix strategies apply to all
three matrix games; game-specific strategies apply to their respective
sequential game.}
\label{tab:opponents}
\small
\begin{tabular}{ll}
\toprule
\textbf{Strategy} & \textbf{Description} \\
\midrule
Always Cooperate     & Unconditionally plays the cooperative action \\
Always Defect        & Unconditionally plays the defective action \\
Tit-for-Tat          & Cooperates first, then mirrors opponent's last move \\
Tit-for-Two-Tats     & Defects only after opponent defects twice in a row \\
Grudger              & Cooperates until opponent defects once, then always defects \\
Pavlov               & Repeats last move if payoffs matched; switches otherwise \\
Suspicious TFT       & Like TFT but defects on the first move \\
Generous TFT         & Like TFT but cooperates with $90\%$ chance after a defection \\
Adaptive             & Cooperates if opponent's cooperation rate exceeds $50\%$ \\
Mixed                & Cooperates with $70\%$ chance each round \\
Random               & Selects uniformly at random from available actions \\
\midrule
Ultimatum Fair       & Offers an equal split; accepts offers above a threshold \\
Ultimatum Low        & Offers a low split; always accepts \\
Trust Fair           & Invests fully; returns one-third of tripled amount \\
Trust Generous       & Invests fully; returns half of tripled amount \\
Public Goods Fair    & Contributes half of endowment \\
Public Goods Free Rider & Contributes a minimal amount \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Episode Structure}

Each episode proceeds as follows:
\begin{enumerate}[nosep]
    \item The environment is initialized with a game and opponent
          strategy via \texttt{reset(game, strategy)}.
    \item Each round, the agent receives a \texttt{GameObservation}
          containing the game description, available actions, round
          history, and cumulative scores.
    \item The agent returns a \texttt{GameAction}; the opponent's
          response is computed internally by the strategy module.
    \item Payoffs are resolved via the game's payoff function and
          appended to the history.
    \item After the final round, \texttt{done=True} is set and
          episode-level statistics are available for metric computation.
\end{enumerate}
